---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# QuAnTeTrack

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/QuAnTeTrack)](https://CRAN.R-project.org/package=QuAnTeTrack)
[![R-CMD-check](https://github.com/MacroFunUV/QuAnTeTrack/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/MacroFunUV/QuAnTeTrack/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Studying fossil vertebrate trackways is crucial as they provide valuable insights into the behavior, locomotion, and environmental interactions of extinct species, offering a dynamic perspective on prehistoric life that skeletal remains alone cannot. Here, I present **QuAnTeTrack**, an integrated tool designed to **semi-automatically extract palaeobiological information** by digitizing footprint and trackway coordinates and analyzing them within the R environment. 

QuAnTeTrack includes functions to read footprint coordinates and trackways stored as **.TPS files**, transforming them into **"track" objects**. These objects can be used to **calculate and plot key parameters** such as **turning angles, trackway distances, step lengths, sinuosity, and straightness**. Additional functions enable the **quantification of velocity** and the testing of scenarios involving **acceleration, deceleration, or stable velocity**, as well as comparing velocity differences between tracks. 

The tool also supports **visualizing movement patterns**, plotting tracks according to velocity. Furthermore, **QuAnTeTrack facilitates the testing of hypotheses related to gregarious behavior or predatory events**. For interactions involving multiple tetrapods, such as **group movement or predation**, trackways are expected to covary more significantly and present fewer intersections than those generated by independent events. QuAnTeTrack estimates this by computing **Frechet and Dynamic Time Warping (DTW) trajectory similarity metrics** and **quantifying track intersections**. These metrics are evaluated through **random simulation procedures** to determine how actual trackway similarities differ from randomly generated ones under various scenarios, including **geographical constraints and resource directionality**.

## Installation

You can install the development version of QuAnTeTrack from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MacroFunUV/QuAnTeTrack")
```

## Example

```{r, echo=TRUE, message=FALSE, max.height='100px'}
library(QuAnTeTrack)
```

Example 1: Paluxy River Dataset

```{r, echo=TRUE, message=FALSE, max.height='100px'}
tpsPaluxyRiver <- system.file("extdata", "PaluxyRiver.tps", package = "QuAnTeTrack")
PaluxyRiver <- tps_to_track(tpsPaluxyRiver, scale = 0.004341493, missing = FALSE, NAs = NULL)
plot_track(PaluxyRiver)
```

Display track parameters within a scrollable box
```{r, echo=TRUE, message=FALSE, max.height='100px'}params_paluxy <- track_param(PaluxyRiver)
print(params_paluxy)
```

```{r, echo=TRUE, message=FALSE, max.height='100px'}
plot_direction(PaluxyRiver, plot_type = "polar_steps")
```

```{r, echo=TRUE, message=FALSE, max.height='100px'}
H_paluxyriver <- c(3.472, 2.200)
Method_paluxyriver <- c("A", "B")
velocity_paluxyriver <- velocity_track(PaluxyRiver, H = H_paluxyriver, method = Method_paluxyriver)
```

```{r, echo=TRUE, message=FALSE, max.height='100px'}
print(velocity_paluxyriver)
```








# Example 2: Mount Tom Dataset

tpsMountTom <- system.file("extdata", "MountTom.tps", package = "QuAnTeTrack")
NAs <- matrix(c(7, 3), nrow = 1, ncol = 2)
MountTom <- tps_to_track(tpsMountTom, scale = 0.004411765, missing = TRUE, NAs = NAs,
                         R.L.side = c("R", "L", "L", "L", "R", "L", "R", "R", "L", "L", 
                                      "L", "L", "L", "R", "R", "L", "R", "R", "L", "R", 
                                      "R", "R", "R"))
plot_track(MountTom)

# Display track parameters within a scrollable box
params_mounttom <- track_param(MountTom)
print(params_mounttom)

plot_direction(MountTom, plot_type = "polar_steps")

H_mounttom <- c(1.380, 1.404, 1.320, 1.736, 1.364, 1.432, 1.508, 1.768, 1.600, 1.848, 
                1.532, 1.532, 0.760, 1.532, 1.688, 1.620, 0.636, 1.784, 1.676, 1.872, 
                1.648, 1.760, 1.612)
velocity_mounttom <- velocity_track(MountTom, H = H_mounttom)
print(velocity_mounttom)

```

